Evaluating `Graphical Perception' with CNNs

### 2018-03-28 PAPER/00_introduction.tex
Editing intro

%Artificial intelligence has taken the world of technology by storm. 
Deep multilayer neural networks are being successfully applied in a wide range of applications that are regularly outperforming humans in object recognition \cite{krizhevsky_imagenet2012, simonyan_very_deep2014,szegedy2015}. 
Originally inspired by neuroscientific discoveries, the recent advances in deep learning have been the direct results of engineering efforts, more specifically in convolutional neural networks (CNNs). 
Our current knowledge of biological vision suggests that CNNs indeed mimic feature learning similar to the early layers of the visual cortex but without incorporating the many details of biological neural networks \cite{yamins2016using, hassabis2017neuroscience, human_vs_machine_vision}.
While there has been significant advancement, fully understanding how CNNs work is a still ongoing process which does not detract from their success~\cite{goodfellow_book, deeplearning_blackbox2017} and recent works target the systematic evaluation of perceptional limits of CNNs~\cite{clevr, not_so_clevr}.

We are interested in applying neural networks to graphs, charts, and visual encodings. A first step in this direction is to evaluate if and how convolutional neural networks perceive low-level graphical elements, the building blocks of information visualizations. This is a very different task than applying such networks to natural images (like the majority of computer vision research) since such scenes are usually more complex and identifiers such as edges are not as prominent. More related towards our goal are the \emph{graphical perception} experiments of Cleveland and McGill~\cite{cleveland_mcgill}. In this paper, we explore whether these experiments which were performed with humans can be reproduced with CNNs. Cleveland and McGill's work in the 1980s has led to many insights for modern information visualization research such as the identification of elementary perceptional tasks or that bar charts are easier to understand than pie charts.

>>>>

Convolutional neural networks (CNNs) have been successfully applied to a wide range of visual tasks, most famously to natural image object recognition~\cite{krizhevsky_imagenet2012, simonyan_very_deep2014, szegedy2015}, for which some claim equivalent or better than human performance. This performance comparison is often motivated by the idea that CNNs model or reproduce the early layers of the visual cortex, even though they do not incorporate many details of biological neural networks or model higher-level abstract or symbolic reasoning~\cite{yamins2016using, hassabis2017neuroscience, human_vs_machine_vision}. While CNN techniques were originally inspired by neuroscientific discoveries, recent advances in processing larger datasets with deeper networks have been the direct results of engineering efforts. Throughout this significant advancement, researchers have aimed to understand why and how CNNs produce such high performance~\cite{goodfellow_book, deeplearning_blackbox2017}, with recent works targetting the systematic evaluation of the visual perception limits of CNNs~\cite{clevr, not_so_clevr}.

One fundamental application of human vision is in understanding data visualizations. This is a task unlike processing natural images, where real-world objects and their effects are abstracted into data and represented with visual marks. As a field, visualization catalogues and evaluates human perception of these figures, such as in the seminal \emph{graphical perception} experiments of Cleveland and McGill~\cite{cleveland_mcgill}. This work describes nine elementary perceptual reasoning tasks, such as position relative to a scale, length, angle, area, and shading density, plus orders their reasoning difficulty. But, with more research interested in the machine analysis of graphs, charts, and visual encodings, it seems pertinent to question whether CNNs are able to process these basic graphical elements, and derive useful measurements from the building blocks of information visualization.

As such, we explore whether these experiments which were performed with humans can be reproduced with CNNs. Cleveland and McGill's work in the 1980s has led to many insights for modern information visualization research such as the identification of elementary perceptional tasks or that bar charts are easier to understand than pie charts.
